\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing income inequality with tax data - Switzerland from 1917 to 2010}

### 2014-07-04 paper/subdocuments/results.Rnw
captions added results.Rnw

<<different_ginis, fig.cap='',fig.pos='H'>>=
df$G_reink[df$G_reink==1]<-NA
df$G_taxed[df$G_taxed==1]<-NA
library(reshape)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
dflong <- melt(df[,c("G_steink","G_reink","G_taxed","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Income"
names(dflong)[4]<-"Gini"
levels(dflong$Income)<-c("taxable income","net income","taxable income after federal tax")
ggplot(dflong[dflong$kanton=="CH",], aes(x=steuerperiode,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=dflong[dflong$kanton=="CH",],aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
#ggplot(dflong, aes(x=steuerperiode,y=Gini,colour=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")
@

@@

<<equivalencescale, fig.cap='',fig.pos='H'>>=
number_ticks <- function(n) {function(limits) pretty(limits, n)}
ggplot(bdlong, aes(x=Jahr,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1970,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=bdlong,aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
@

@@

<<reldist20032010, echo=FALSE, fig.cap='',fig.pos='H'>>=

@@

<<specialcases9394, fig.cap='',fig.pos='H'>>=
#plot relative distribution

@@

<<specialcases10, fig.cap='',fig.pos='H'>>=

@@

%<<with_without_zeros_Gini,warning=FALSE, fig.cap='',fig.pos='H'>>=
%df <- read.csv("../../data/ginis_und_perzentile_normal.csv",sep="\t")
%#library(reshape)
%dflong <- melt(df[,c("G_steink","G_steink0","G_steink_halb","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
%names(dflong)[3]<-"Type"
%names(dflong)[4]<-"Gini"
%levels(dflong$Type)<-c("without zeros","including zeros","imputed values for zeros")
%ggplot(dflong[dflong$kanton=="CH" & dflong$steuerperiode>=1995.5,], 
     %  aes(x=steuerperiode,y=Gini,shape=Type,linetype=Type))+
 % scale_y_continuous(limits=c(0.18,0.6))+
 % scale_x_continuous(limits=c(1994,2012),breaks=number_ticks(10)) +
 % geom_line(size=0.5)+
 % geom_point(aes(shape=Type),size=3)+
 % theme_bw()+
 % xlab("tax period")+
 % theme(legend.justification=c(0,1),legend.position=c(0, 1))

%# Kantonale Unterschiede
%#####
%#ggplot(dflong[dflong$steuerperiode>=1995.5,], aes(x=steuerperiode,y=Gini,shape=Type,linetype=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")+theme(axis.text.x = element_text(angle = 60, hjust = 1))+theme(legend.justification=c(1,0),legend.position=c(1, 0))
%#####
%@

@@

%par(mfrow=c(1,1))
@

@@

<<BruelhartvsHABE,warning=FALSE, fig.cap='',fig.pos='H'>>=
#library(ggplot2)
#library(reldist, quietly=TRUE, warn.conflicts = FALSE)

@@

# Select married - HABE

@@

gA0 <- reldist(y=habe10.married$inc_habe, yo=bd.2010.ind,
               ywgt=habe10.married$gew,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.0), ylab="",
               xlab="Proportion of FTA Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))

#################
# Separate comparison for unmaried

### Not corrected!!!

@@

From the discussion in the data section we would expect differences between the income distributions from survey and tax data. Within the FTA data we observe zeros for incomes below the threshold to be liable for federal tax while survey data might cover this range. On the other hand we expect underreporting from both lowest percentiles and highest percentiles (middle class bias) within the survey data. Figure \ref{fig:BruelhartvsHABE1} (left) however reveals a more critical issue related to tax data, that is the median location of income compared to survey data. Although we try to measure similar concepts of income, survey data shows a median (93.000 CHF) more than twice as big as tax data (44.600 CHF). The issue here is clearly the assumptions of household composition. While survey data is likely to capture the correct household composition, tax data can only approximate households by using marital status. Splitting the data in married and unmarried tax units underpins this argument. From figure \ref{fig:BruelhartvsHABE2} we can see that married tax units (FTA data) and household with married couples (survey data) are better (but still not perfectly) comparable. Figure \ref{fig:BruelhartvsHABE2} (right) shows the expected shape difference between the two distributions:  survey data has a bias towards the (median-adjusted) 80\% to 90\% percentile (of the FTA data distribution). Top percentiles are badly covered by survey data, suggesting that inequality measures based on survey data might underestimate inequality development that arises from changes in the incomes of the rich. Though, survey data can be of interest if one is interested in the lower 20\% of the income distribution.

@@

<<benplot, fig.cap='',fig.pos='H'>>=
 ### libs
#library(reshape, quietly=TRUE, warn.conflicts = FALSE)

>>>>

<<different_ginis, fig.cap='Gini coefficients for different income measures',fig.pos='H'>>=
df$G_reink[df$G_reink==1]<-NA
df$G_taxed[df$G_taxed==1]<-NA
library(reshape)
number_ticks <- function(n) {function(limits) pretty(limits, n)}
dflong <- melt(df[,c("G_steink","G_reink","G_taxed","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Income"
names(dflong)[4]<-"Gini"
levels(dflong$Income)<-c("taxable income","net income","taxable income after federal tax")
ggplot(dflong[dflong$kanton=="CH",], aes(x=steuerperiode,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1982,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=dflong[dflong$kanton=="CH",],aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
#ggplot(dflong, aes(x=steuerperiode,y=Gini,colour=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")
@

@@

<<equivalencescale, fig.cap='Gini coefficients with and without equivalization',fig.pos='H'>>=
number_ticks <- function(n) {function(limits) pretty(limits, n)}
ggplot(bdlong, aes(x=Jahr,y=Gini,shape=Income))+
  scale_y_continuous(limits=c(0.3,0.55))+
  scale_x_continuous(limits=c(1970,2012),breaks=number_ticks(10)) +
  geom_point(aes(shape=Income),size=3)+
  geom_line(data=bdlong,aes(linetype=Income))+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))
@

@@

<<reldist20032010, echo=FALSE, fig.cap='Distribution change over time',fig.pos='H'>>=

@@

<<specialcases9394, fig.cap='Comparison of normal and special cases 1993/94',fig.pos='H'>>=
#plot relative distribution

@@

<<specialcases10, fig.cap='Comparison of normal and special cases 2010',fig.pos='H'>>=

@@

<<with_without_zeros_Gini,warning=FALSE, fig.cap='Gini coefficients with, without and imputed non-taxed units',fig.pos='H'>>=
df <- read.csv("../../data/ginis_und_perzentile_normal.csv",sep="\t")
#library(reshape)
dflong <- melt(df[,c("G_steink","G_steink0","G_steink_halb","kanton","steuerperiode")], id=c("kanton","steuerperiode"))
names(dflong)[3]<-"Type"names(dflong)[4]<-"Gini"
levels(dflong$Type)<-c("without zeros","including zeros","imputed values for zeros")
ggplot(dflong[dflong$kanton=="CH" & dflong$steuerperiode>=1995.5,], 
       aes(x=steuerperiode,y=Gini,shape=Type,linetype=Type))+
  scale_y_continuous(limits=c(0.18,0.6))+
  scale_x_continuous(limits=c(1994,2012),breaks=number_ticks(10)) +
  geom_line(size=0.5)+
  geom_point(aes(shape=Type),size=3)+
  theme_bw()+
  xlab("tax period")+
  theme(legend.justification=c(0,1),legend.position=c(0, 1))

# Kantonale Unterschiede
#####
#ggplot(dflong[dflong$steuerperiode>=1995.5,], aes(x=steuerperiode,y=Gini,shape=Type,linetype=Type))+geom_line()+facet_wrap(~kanton)+theme_bw()+xlab("tax period")+theme(axis.text.x = element_text(angle = 60, hjust = 1))+theme(legend.justification=c(1,0),legend.position=c(1, 0))
#####
@

@@

%par(mfrow=c(1,1))
%@

@@

<<BruelhartvsHABE,warning=FALSE, fig.cap='Comparison of tax and survey data',fig.pos='H'>>=
#library(ggplot2)
#library(reldist, quietly=TRUE, warn.conflicts = FALSE)

@@

bd.2010.ind<-bd.2010.ind/1000
@

<<BruelhartvsHABEmarried,warning=FALSE, fig.cap='Comparison of tax and survey data for married',fig.pos='H'>>=
# Select married - HABE

@@

gA0 <- reldist(y=habe10.married$inc_habe, yo=bd.2010.ind,
               ywgt=habe10.married$gew,
               smooth=0.4, ci=FALSE,
               show="residual",
               bar=TRUE, quiet=FALSE,
               ylim=c(0.5,2.0), ylab="",
               xlab="Proportion of FTA Distribution")
title(main=paste("Effect of different shape"))
abline(h=1,lty=2)
par(mfrow=c(1,1))
@

<<BruelhartvsHABEunmarried,warning=FALSE, fig.cap='Comparison of tax and survey data for unmarried',fig.pos='H'>>=
#################
# Separate comparison for unmaried

@@

From the discussion in the data section we would expect differences between the income distributions from survey and tax data. Within the FTA data we observe zeros for incomes below the threshold to be liable for federal tax while survey data might cover this range. On the other hand we expect underreporting from both lowest percentiles and highest percentiles (middle class bias) within the survey data. Figure \ref{fig:BruelhartvsHABE} (left) however reveals a more critical issue related to tax data, that is the median location of income compared to survey data. Although we try to measure similar concepts of income, survey data shows a median (93.000 CHF) more than twice as big as tax data (44.600 CHF). The issue here is clearly the assumptions of household composition. While survey data is likely to capture the correct household composition, tax data can only approximate households by using marital status. Splitting the data in married and unmarried tax units underpins this argument. From figure \ref{fig:BruelhartvsHABEmarried} we can see that married tax units (FTA data) and household with married couples (survey data) are better (but still not perfectly) comparable. Figure \ref{fig:BruelhartvsHABEmarried} (right) shows the expected shape difference between the two distributions:  survey data has a bias towards the (median-adjusted) 80\% to 90\% percentile (of the FTA data distribution). Top percentiles are badly covered by survey data, suggesting that inequality measures based on survey data might underestimate inequality development that arises from changes in the incomes of the rich. Though, survey data can be of interest if one is interested in the lower 20\% of the income distribution.

@@

<<benplot, fig.cap='Gini coefficients over time',fig.pos='H'>>=
 ### libs
#library(reshape, quietly=TRUE, warn.conflicts = FALSE)
